\documentclass[12pt]{article}
\title{M374M Homework 1 \\
  \normalsize{\S~1.1 \#4$^1$, 5, 9$^2$, 13, 14$^3$}}
\author{Hershal Bhave (hb6279)}
\date{Due 2016-02-01}

\usepackage{macros}

\begin{document}

\tableofcontents

\section{Dimensional Analysis}
\subsection{Example}
Let $q = \alpha p^2$, where $[\alpha] = \frac{L}{T^2}$, $[p] = T$. Find $[q]$,
$\left[\frac{dq}{dp}\right]$.

\subsection{Definition}
$q$ is \emph{dimensionless} if $[q] = 1$. Angles are dimensionless, for example.

\subsection{Examples}
\subsubsection{}
Let $y = f(x)$, where $[y] = [x]$. Then $\frac{dy}{dx}$ is dimensionless.

\subsubsection{}
By definition, pure numbers are dimensionless

\begin{equation}
  \begin{aligned}
    g &= 10 \text{m}/\text{s}^2 \\
    g + g &= 2g = 20 \text{m}/\text{s}^2 \\
    g \cdot g \cdot g & = g^3 = 1000 \text{m}^3/\text{s}^6
  \end{aligned}
\end{equation}

\subsection{Remark}
Some functions make dimensional sense only if inputs are dimensionless.

\begin{equation}
  \begin{aligned}
    e^q &= 1 + q + \frac{q^2}{2!} + \frac{q^3}{3!} + \cdots \\
    [q] &= 1 \implies [e^q] = 1 \\
    [q] &= L \implies [e^q] = ?
  \end{aligned}
\end{equation}

\subsection{Definition}
An equation in ``standard form'' $F(q, p, \ldots) = 0$ is called unit-free if
each term has the same dimenison.

\subsection{Example}
Galileo's Law.

\begin{equation}
  \begin{aligned}
    x &= \frac{gt^2}{2} \\
    x - \frac{gt^2}{2} &= 0 \\
    F(x,t,g) &= 0
  \end{aligned}
\end{equation}

Where $[x] = L$, $[\frac{gt^2}{2}] = [gt^2] = L$.

$\therefore F(x,t,g)=0$ is unit-free.

\subsection{Remarks}
\begin{enumerate}
\item All equations we consider are unit-free
\item Differentiation, integration preserves unit-free propery of an equation
\end{enumerate}

\begin{equation}
  \begin{aligned}
    \frac{d^2x}{dt^2} &= g \text{unit-free} \\
    \frac{dx}{dt} &= gt + v_0 \\
    x &= \frac{1}{2}gt^2 + v_0 t + x_0 \\
  \end{aligned}
  %% $\therefore$ unit-free ODE $\implies$ unit-free solution.
\end{equation}

\subsection{Definition}
Let $Q = {q_1, ..., q_m}$ be a set of quantities involving a set $D =
{D_1,...,D_n}$ of dimensions so that

\begin{equation}
  \begin{aligned}
    [q_1] &= D_1^{a_{11}} D_2^{a_{21}}\cdots D_n^{a_{n1}} \\
    \vdots & \\
    [q_2] &= D_1^{a_{21}} D_2^{a_{22}}\cdots D_n^{a_{n2}} \\
  \end{aligned}
\end{equation}
for some exponents $a_{11}, \ldots, a_{nm}$.

By the dimension matrix for $Q$, $D$, we mean
\begin{equation}
  A =
  \begin{pmatrix}
    a11 & a12 & \cdots & a1m \\
    a21 & a22 & \cdots & a2m \\
    \vdots & & & \\
    an1 & an2 & \cdots & anm \\
  \end{pmatrix}
  \in \mathbb{R}^{n\times m}
\end{equation}

\subsection{Example}
$m = 3 | Q = {x, t, g}$, $n = 2 | D = {L, T}$.
\begin{equation}
  \begin{aligned}
    [x] &= L^1 T^0 \\
    [t] &= L^0 T^1 \\
    [g] &= L^1 T^{-2}
  \end{aligned}
\end{equation}

\begin{equation}
  A =
  \begin{pmatrix}
    1 & 0 & 1 \\
    0 & 1 & -2 \\
  \end{pmatrix}
  \in \mathbb{R}^{2\times3}
\end{equation}

\subsection{Remarks}
All dimensionless combinations of $q_1,\ldots,q_m$ can be found from the null
vectors of $A$. That is,

$$Z = q_1^{v_1}, q_2^{v_2}, \cdots, q_n^{v_n}$$ where $v_1, \ldots, v_m$
arbitrary.

Then
\begin{equation}
  \begin{aligned}
    [Z] &= 1 \rightleftarrows Av=0, \\
    v &= \begin{pmatrix} v_1 \\ \vdots \\ v_m \end{pmatrix}
  \end{aligned}
\end{equation}

\subsection{Example}
Find all dimensionless combinations of $x, t, g$.

\begin{equation}
  \begin{aligned}
    Z &= x^{v_1}t^{v_2}g^{v_3} \\
    [Z] &= [x]^{v_1}[t]^{v_2}[g]^{v_3} \\
    &= L^{v_1}T^{v_2}(LT^{-2})^{v_3} \\
    &= L^{v_1+v_3} T^{v_2-2v_3} \\
    \implies [Z] = 1 & \rightleftarrows  v_1+v_3=0, v_2-2v^3=0
  \end{aligned}
\end{equation}

\begin{equation}
  \begin{aligned}
    \begin{pmatrix}
      1 & 0 & 1 \\
      0 & 1 & -2
    \end{pmatrix}
    \begin{pmatrix}
      v_1 \\ v_2 \\ v_3 \\
    \end{pmatrix}
    &=
    \begin{pmatrix}
      0 \\ 0
    \end{pmatrix}
    \implies Av &= 0
  \end{aligned}
\end{equation}

\begin{equation}
  v = c
  \begin{pmatrix}
    -1 & 2 & 1
  \end{pmatrix}
\end{equation}

  Where $c$ is free.

\begin{equation}
  \begin{aligned}
    c &= 1 \rightarrow v =
    \begin{pmatrix}
      -1 \\ 2 \\ 1
    \end{pmatrix} \rightarrow
    Z = x^{-1} t^2 g^1 = \frac{t^2g}{x} \\
    c &= 2 \rightarrow v=
    \begin{pmatrix}
      -2 \\ 4 \\ 2
    \end{pmatrix} \rightarrow
    Z = \frac{t^4g^1}{x^2}
  \end{aligned}
\end{equation}

The cases where $c=1$ and $c=2$ are the same combination, so we don't have
independent results.

\newpage
\section{Characteristic Scales}
\subsection{Definition}
Non-zero constants $t_c$, $y_c$ are called characteristic scales for a function
$y=f(t)$ if the following is true:

\begin{enumerate}
\item \begin{itemize}
\item $[t_c] = [t]$
\item $[y_c] = [t]$
\end{itemize}

\item Main features of $y=f(t)$ graph are clearly visible in a $mt_c \times ny_c$
   window for moderate values of $m,n$ ($1 \le m, \; n \le 10$).
\end{enumerate}

\subsection{Examples}
\subsubsection{Example 1}

$y=A\sin(\pi t/b), t \ge 0$, $A$, $b$ constants, $A = 1 \text{ meter}$, $b=1
\text{ second}$. Reference fig. 1.1.

Characteristic scales are $t_c=b, y_c=A$ since $2b \times 2A$ window captures
main features of graph.

Can understand the features of the graph within this window. Reference fig. 1.2.

Windows of significantly different sizes would give poor representations of
functions. Reference fig. 1.3.

\subsubsection{Example 2}

$y = Ae^{-t/b}, t\ge0$ where $[y]=L$, $[t] = T$, $A$, $b$ constants.

Reference fig. 1.4. Characteristic scales are $t_c=b$, $y_c=A$ since window
$10b \times A$ captures main features.

\subsubsection{Example 3}

Fig. 1.5 has two sets of characteristic scales; two window sizes are needed to
represent the features, as shown in figs. 1.6 and 1.7.

\subsection{Procedure}
Given some (potentially indirect) definition of a graph, how do we indentify the
scales?

If $y=f(t)$ is defined by an equation involving constants $g_1, \ldots, g_m$,
then characteristic scales $t_c$, $y_c$ can be found by solving:

\begin{equation}
  \begin{aligned}
    t_c &= g_1^{\alpha_1}, \ldots, g_m^{\alpha_m} \\
    y_c &= g_1^{\beta_1}, \ldots, g_m^{\beta_m}
  \end{aligned}
\end{equation}

for $\alpha_1, \ldots, \alpha_m$ and $\beta_1, \ldots, \beta_m$.

\subsection{Examples}
\subsubsection{Example 1}

Find characteristic scales $t_c$, $y_c$ for $y=f(t)$ defined by

\begin{equation}
  \left\{
  \begin{aligned}
    \frac{dy}{dt} &= \eta y, \quad t\ge0 \\
    y_{(0)} &= \lambda \\
  \end{aligned} \right.
\end{equation}

$[y] = \Eta , [t]=T$.

Dimensions of $\eta$, $\lambda$:

\begin{equation}
  \begin{aligned}
    \frac{\Eta}{T} &= [\eta]\Eta \rightarrow [\eta]=\frac{1}{T} \\
    \Eta &= [\lambda]
  \end{aligned}
\end{equation}

Scales for $t$, $y$:

\begin{equation}
  \begin{aligned}
    t_c &= \eta^{\alpha_1} \lambda^{\alpha_2} \\
    \implies T &= T^{-\alpha_1}\Eta^{\alpha_2} \\
    \implies \alpha_1 &= -1,\quad \alpha_2 = 0 \\
    \implies t_c &= \frac{1}{\eta}\\
  \end{aligned}
\end{equation}

\begin{equation}
  \begin{aligned}
    y_c &= \eta^{\beta_1} \lambda^{\beta_2} \\
    \implies \Eta &= T^{-\beta_1}\Eta^{\beta_2} \\
    \implies \beta_1 &= 0,\quad \beta_2 = 1 \\
    \implies y_c &= \lambda
  \end{aligned}
\end{equation}

\subsubsection{Example 2}

Do the same for
\begin{equation}
  \left\{
  \begin{aligned}
    \frac{d^2y}{dt^2} &= \eta \frac{dy}{dt} + \mu y^2,\quad t \ge 0 \\
    y_{(0)} &= \lambda, \quad \frac{dy}{dt}(0) = 0
  \end{aligned} \right.
\end{equation}
$[y]=L$, $[t]=T$, $\eta$, $\mu$, $\lambda$ constants.

Dimensions of $\eta$, $\mu$, $\lambda$:
\begin{equation}
  \begin{aligned}
    [\lambda] &= L \\
    [\eta]\frac{L}{T} &= \frac{L}{T^2} \quad\longrightarrow\quad [\eta] = \frac{1}{T} \\
    [\mu]L^2 &= \frac{L}{T^2} \quad\longrightarrow\quad [\mu] = \frac{1}{LT^2} \\
  \end{aligned}
\end{equation}

Scale for $t$:
\begin{equation}
  \begin{aligned}
    t_c &= \eta^{\alpha_1}\mu^{\alpha_2}\lambda^{\alpha_3} \\
    &= T^{-\alpha_1}(L^{-1}T^{-2})^{\alpha_2}L^{\alpha_3} \\
    \implies L^0T^1 &= L^{-\alpha_2+\alpha_3}T^{\alpha_1-2\alpha_2} \\
  \end{aligned}
\end{equation}

\begin{equation}
  \begin{pmatrix}
    -1 & -2 & 0 \\
    0 & -1 & 1 \\
  \end{pmatrix}
\begin{pmatrix}
  \alpha_1 \\
  \alpha_2 \\
  \alpha_3 \\
\end{pmatrix}
=
\begin{pmatrix}
  1\\0
\end{pmatrix}
\end{equation}

Where $\alpha_3$ is free.

\begin{equation}
  \begin{aligned}
    -\alpha_1-2\alpha_2 &= 1 \rightarrow \alpha_1=-2\alpha_2-1=-2\alpha_3-1\\
    -\alpha_2 + \alpha_3 = 0 \rightarrow \alpha_2&=\alpha_3
  \end{aligned}
\end{equation}

\begin{equation}
  \begin{pmatrix}
    \alpha_1 \\ \alpha_2 \\\alpha_3 \\
  \end{pmatrix}
  =
  \begin{pmatrix}
    -2\alpha_3-1 \\ \alpha_3 \\\alpha_3 \\
  \end{pmatrix}
  = \alpha_3
  \begin{pmatrix}
    -2 \\ 1 \\ 1
  \end{pmatrix}
  +
  \begin{pmatrix}
    -1 \\ 0 \\ 0
  \end{pmatrix}
\end{equation}

Where $\alpha_3$ is free. $\alpha_3=0, \quad \alpha_3\ne0$ gives two scales for
t\_c:

\begin{equation}
  \alpha_3=0, \quad
  \begin{pmatrix}
    -1 \\ 0 \\ 0
  \end{pmatrix}
\end{equation}

\begin{equation}
  \implies t_c = \frac{1}{\eta}.
\end{equation}

And

\begin{equation}
  \alpha_3 = \frac{-1}{2}, \quad
  \begin{pmatrix}
    \alpha_1 \\ \alpha_2 \\ \alpha_3 \\
  \end{pmatrix}
  =
  \begin{pmatrix}
    0 \\ -1/2 \\ -1/2
  \end{pmatrix}
\end{equation}

\begin{equation}
  \implies t_c = \frac{1}{\sqrt{\mu\lambda}}
\end{equation}

Same for $y_c$

\begin{equation}
  \begin{aligned}
    y_c &= \frac{\eta^2}{\mu} \\
    y_c &= \lambda
  \end{aligned}
\end{equation}

\newpage
\section{Dimensionless Forms}
\subsection{Definition}
Consider an equation involving vars $v_1,\ldots,v_k$. By the dimensionless form
of an equation with respect to scales $v_1^c,\ldots,v_k^c$ we mean the equation
obtained by the substitution $\overline{v_i}=v_i/v_i^c,\quad i=1,\ldots,k$ where
$\overline{v_i}$ is a dimensionless variable.

\subsection{Example}
\begin{equation}
  \left\{
  \begin{aligned}
    \frac{dy}{dt} &= \eta y+r, \quad t\ge0 \\
    y_{(0)} &= \lambda \\
  \end{aligned}
  \right.
\end{equation}

\begin{equation}
  [y] = \Theta,\quad[t]=T,\quad \eta,\lambda,r > 0 \text{ constants}
\end{equation}

Find dimensionless equation with respect to scales
$t_c=\frac{1}{\eta},\quad y_c=\lambda$.

Dimensionless vars
\begin{equation}
  \begin{aligned}
    \bar{y} &= y/y_c = y/\lambda,\quad & \bar{t} &= t/t_c = \eta t \\
    y &= \lambda \bar{y}, \quad & t&=\bar{t}/\eta \\
  \end{aligned}
\end{equation}

Derivitive relation
\begin{equation}
  \begin{aligned}
    \frac{d\bar{y}}{d\bar{t}} &= \frac{d\bar{y}}{dt} \cdot \frac{dt}{d\bar{t}}
  \end{aligned}
\end{equation}

Where $\frac{d\bar{y}}{dt} = \frac{d}{dt}(y/\lambda) = \frac{1}{\lambda}\frac{dy}{dt}$ and
$\frac{dt}{d\bar{t}} = \frac{1}{\eta}$. Thus

\begin{equation}
  \begin{aligned}
    \frac{d\bar{y}}{d\bar{t}} &= \frac{1}{\eta\lambda}\frac{dy}{dt} \\
    \therefore \frac{dy}{dt} &= \eta\lambda\frac{d\bar{y}}{d\bar{t}} \\
  \end{aligned}
\end{equation}

In general, $\frac{dy}{dt} = \frac{y_c}{t_c} \cdot \frac{d\bar{y}}{d\bar{t}}$.
Substituted into equations,

\begin{equation}
  \begin{aligned}
    \frac{dy}{dt} &= \eta y+r, &\quad t&\ge0 \\
    \eta\lambda\frac{d\bar{y}}{d\bar{t}} &= \eta\lambda\bar{y}+r, &\quad \bar{t}/\eta&\ge0 \\
    \frac{d\bar{y}}{d\bar{t}} &= \bar{y}+\frac{r}{\eta\lambda}, &\quad \bar{t}&\ge0 \\
    y &= \lambda, &\quad t&=0 \\
    \lambda\bar{y} &= \lambda, &\quad \bar{t}/\eta &= 0 \\
    \bar{y} &= 1, &\quad\bar{t}&=0. \\
  \end{aligned}
\end{equation}

We end up with dimensionless equations
\begin{equation} \left\{
  \begin{aligned}
    \frac{d\bar{y}}{d\bar{t}} &= \bar{y} + \bar{r}, &\quad \bar{t}&\ge0 \\
    \bar{y} &= 1, &\quad \bar{t} &= 0
  \end{aligned} \right.
\end{equation}

Where $\bar{r} = \frac{r}{\eta\lambda}$ is a dimensionless constant.

\subsection{Remarks}
\begin{enumerate}
\item Different choices of scales lead to different dimensionless equations
  \begin{equation}
    \begin{aligned}
      t_c &= \lambda/r \\
      y_c &= \lambda
    \end{aligned}
  \end{equation}
  Which leads to

  \begin{equation}
    \begin{aligned}
      \bar{t} &= t/t_c = rt/\lambda \\
      \bar{y} &= y/y_c = y/\lambda \\
    \end{aligned}
  \end{equation}


  Which leads to
  \begin{equation}
    \begin{aligned}
      \frac{d\bar{y}}{d\bar{t}} &= \bar{\eta}\bar{y} + 1, &\quad \bar{t}&\ge0 \\
      \bar{y} &= 1 &\quad \bar{t} &= 0 \\
      \bar{\eta} &= \frac{\eta\lambda}{r} \\
    \end{aligned}
  \end{equation}

  Where $\bar{y}$ and $\bar{\eta}$ are dimensionless.

\item Solutions of dimensionful and dimensionless equations are equivalent

  \begin{equation}
    \text{dimensioned solution}\quad\xleftrightarrow[\bar{t} = t/t_c]{\bar{y} = y/y_c}\quad
    \text{dimensionless solution}
  \end{equation}

\item Dimensionless equations allow comparisons to be made
\end{enumerate}

For example, we may not compare $a$, $b$, $c$, $d$ in \cref{eq:dim-full-eq}
since dimensions are different. By comparison, we may compare $\bar{a}$,
$\bar{b}$, $\bar{c}$, $\bar{d}$ in \cref{eq:dim-less-eq} since all variables are
dimensionless.

\begin{equation}
  \label{eq:dim-full-eq}
  \frac{d^2u}{dt^2} = au + bu^2 + ce^{-u/d}
\end{equation}

\begin{equation}
  \label{eq:dim-less-eq}
  \frac{d^2\bar{u}}{d\bar{t^2}} = \bar{a} \bar{u} + \bar{b} \bar{u^2} + \bar{c}e^{-\bar{u}/\bar{d}}
\end{equation}

\subsection{Example}
Vertical motion of a ball can be described using the variables in
\cref{fig:ball-model} and the equation

\begin{figure}
  \centering
  \begin{tabularx}{0.5\textwidth}{XX}
    variable & quantity \\ \hline
    $g$ & gravity \\
    $\eta$ & air resistance \\
  \end{tabularx}
  \caption{vertical motion of a ball model}
  \label{fig:ball-model}
\end{figure}

\begin{equation}
  \frac{d^2x}{dt^2} = -g - \eta\frac{dx}{dt}, \quad t\ge0
\end{equation}

Where
\begin{equation}
  x=0,\quad \frac{dx}{dt} = v_0, \quad t=0
\end{equation}

Where $\eta$, $g$, $v_0$ are constants greater than zero.

Under what condition on $\eta$ do we expect air effects to be small?

First, choose scales. In the limiting case of no air resistance ($\eta=0$), the
solution is determined by $g$, $v_0$. Using these constants, we get
$t_c=v_0/g, x_c=v_0^2/g$. Define dimensionless variables
$$\bar{x}=x/x_c,\quad \bar{t}=t/t_c$$ and dimensionless
equations
\begin{equation}
  \frac{d^2\bar{x}}{d\bar{t^2}} = -\bar{g} -
  \bar{\eta}\frac{d\bar{x}}{d\bar{t}}, \quad \bar{t}\ge0
\end{equation}

Where dimensionless variables
\begin{equation}
  \begin{aligned}
    \bar{x} &= 0,\quad \frac{d\bar{x}}{d\bar{t}}&=1 \text{ at } \bar{t}=0 \\
    \bar{g} &= 1, \quad \bar{\eta} &= \frac{\eta v_0}{g} \\
  \end{aligned}
\end{equation}

Expect air effects to be small (when solution is viewed in window with sales
$t_c$, $x_c$) where $\bar{\eta} << \bar{g}$, $\frac{\eta v_0}{g} << 1$, and
$\eta << \frac{g}{v_0}$.

\end{document}
