\item[Stability of Equilibria of Nonlinear Systems in 2D] \hfill \\
  \begin{enumerate}
  \item Determine equilibria
    \begin{itemize}
    \item Direct Substitution
    \item Coupled
    \item Uncoupled
    \end{itemize}
  \item Linearize solution by Jacobian:
    \begin{equation*}
      J = \begin{pmatrix}
        f_x & f_y \\ g_x & g_y \\
      \end{pmatrix}
    \end{equation*}
  \item Eigenvalues of Jacobian about the equilibria characterize the respective
    equilibria. Let $J_*=J(x_*,y_*)$; obtain eigenvalues by solving
    $\lambda^2-\tr{J_*}+\det{J_*}$ or let $D={(\tr J_*)}^2-4\det J_*$ and
    reference the table. Equilibrium is a node if $D\ge0$; spiral otherwise.
    \begin{itemize}
    \item $\det J(x_*,y_*)<0 \implies$ saddle point
    \item $\det J(x_*,y_*)>0, \tr J(x_*,y_*)>0 \implies$ asy.\ unstable
    \item $\det J(x_*,y_*)>0, \tr J(x_*,y_*)<0 \implies$ asy.\ stable
    \item $\det J(x_*,y_*)>0, \tr J(x_*,y_*)=0 \implies$ center point
    \end{itemize}
  \end{enumerate}

\item[SIR Model] Plain SIR
  is $$\dot{S}=-aSI,\quad\dot{I}=aSI-bI,\quad\dot{R}=bI.$$ SIR with
  re-susceptibility and vaccination is
  $$\dot{S}=-aSI+\mu R-vS,\quad\dot{I}=aSI-bI,\quad\dot{R}=bI-\mu R-vS.$$
  Typically don't use $R$ since $R=N-S-I$ and omitting $R$ yields a 2D system.

\item[Two-Dimensional Nonlinear Systems] \hfill\\
  \begin{enumerate}
  \item Find Nullclines
    \begin{enumerate}
    \item Find $f(x,y)=0$ and $g(x,y)=0$ in isolation
    \end{enumerate}
  \item Find Equilibria
    \begin{enumerate}
    \item This is when the nullclines cross.
    \item Find zeros using $f$ and $g$ equations together
    \end{enumerate}
  \item Identify Equilibria Stability
    \begin{enumerate}
    \item Linearize system if not linear
    \item Equilibrium is a node if eigenvalues are real; spiral if complex.
      \begin{itemize}
      \item $\Re(\lambda_1) < \Re(\lambda_2) < 0 \implies$ asy.\ stable
      \item $\Re(\lambda_1) > \Re(\lambda_2) > 0 \implies$ asy.\ unstable
      \item $\Re(\lambda_1) < 0 < \Re(\lambda_2) \implies$ saddle
      \item $\Re(\lambda_1) \equiv \Re(\lambda_2) \equiv 0$ no conclusion; use
        other method (Lyanpunov)
      \end{itemize}
    \end{enumerate}
  \end{enumerate}

\item[Two-Dimensional Linear Systems] \hfill \\
  \begin{enumerate}
  \item Find equilibria. If $\det A\ne0$, then equilibrium is null space of
    coefficient matrix. Otherwise $(0,0)$ is only equilibrium.
  \item Find eigenvalues $\lambda_1$, $\lambda_2$ to characterize solution
  \item Find eigenvectors $\hat{u_1}$, $\hat{u_2}$ to draw nullclines
  \item General solution is in the form $v(t)=C_1e^{\lambda_1
      t}\hat{u_1}+C_2e^{\lambda_2 t}\hat{u_2}$. If eigenvalues are complex use
    Euler's formula where $\lambda_n=\alpha\pm i\beta$ and
    $\hat{u_n}=\hat{\gamma}\pm i\hat{\eta}$: General solution is $v(t) =
    C_1e^{\alpha t}[(\cos\beta t)\hat{\gamma} - (\sin\beta t)\hat{\eta}] +
    C_2e^{\alpha t}[(\sin\beta t)\hat{\gamma} - (\cos\beta t)\hat{\eta}]$.
  \end{enumerate}

\item[Perturbation Methods] A perturbed equation is regular if
  $\text{(Eq)}_{\epsilon=0}$ is of same type as $\text{(Eq)}_{\epsilon>0}$.

\item[Regularly Perturbed w/ Algebraic Solutions] \hfill \\
  \begin{enumerate}
  \item Pick an $\alpha>0$ and let $x(\epsilon) = x_0+\epsilon^{\alpha}x_1 +
    \epsilon^{2\alpha}x_2 + \cdots$.
  \item Substitute into the original equation. Taylor expand nonlinear
    expresions to obtain a polynomial approximation in terms of $\epsilon$.
    Remember to use the chain and product rule and derive in terms of
    $\epsilon$.
    E.g.\ for $f(\epsilon)={x(\epsilon)}^2$,
    \begin{align*}
      f(\epsilon)
      &= f(0) + \epsilon f'(0) + \frac{\epsilon^2}{2!}f''(0)+\cdots \\
      &= x^2_{0} + \epsilon{[2x(\epsilon)x'(\epsilon)]}_{\epsilon=0} +
        \frac{\epsilon^2}{2}{[2x'(\epsilon)x'(\epsilon)+2x(\epsilon)x''(\epsilon)]}_{\epsilon=0}
        + \cdots \\
      &= x_0^2+\epsilon[2x_0x_1] + \frac{\epsilon^2}{2}[2x_1^2 + 2x_0 2x_2] + \cdots \\
    \end{align*}
  \item Collect powers of $\epsilon$
  \item Solve each independent $\epsilon^n$ equation.
  \end{enumerate}

\item[Regularly Perturbed w/ Differential Solutions] \hfill \\
  \begin{enumerate}
  \item Pick an $\alpha>0$ and let $x(t,\epsilon) =
    x_0(t)+\epsilon^{\alpha}x_1(t) + \epsilon^{2\alpha}x_2(t) + \cdots$.
  \item Substitute into the original equation (IVP + ICs). Taylor expand
    nonlinear expresions to obtain a polynomial approximation in terms of
    $\epsilon$. Remember to use the chain and product rule and derive in terms
    of $\epsilon$. E.g.\ for $f(t,\epsilon)=x{(t,\epsilon)}^3$
    \begin{align*}
      f(t,\epsilon)
      &= f(t,0) + \epsilon f(t,0) + \frac{\epsilon}{2!}f''(t,0) + \cdots \\
      &= {[x^3(t,\epsilon)]}_{\epsilon=0} + \epsilon {[3{x(t,\epsilon)}^2x'(t,\epsilon)]}_{\epsilon=0} + \cdots \\
      &= x^3_0(t) + \epsilon[3x_0^2(t)x_1(t)] + \cdots \\
    \end{align*}
  \item Collect powers of $\epsilon$
  \item Solve each independent $\epsilon^n$ ODE.\@
  \end{enumerate}

\item[Taylor Series]
  $$f(x)=f(a)+f'(a)(x-a)+\frac{f''(a)}{2!}{(x-a)}^2+\frac{f^{(3)}(a)}{3!}{(x-a)}^3+\cdots$$
