\item[Euler Equation]
  $$L_y-\od{}{x}L_{y'}+\od[2]{}{x}L_{y''}-\cdots+{(-1)}^n\od[n]{}{n}L_{y^{(n)}}=0$$
  $$L_y(y(x),y'(x)) - y'(x)L_{y'}(x,y(x),y'(x))=c$$
  $$L_{y'}(x,y')=c$$
\item[1st Order Fixed-Fixed] Given
  $$F(y) = \int_a^b f(x)\dd{x}, \quad y(a)=x_0, \quad y(b)=x_1$$
  \begin{enumerate}
  \item Let
    \begin{align*}
      L(x,y(x),y'(x)) &= f(x) \\
      L_{y}(x,y(x),y'(x)) &= \od{f(x)}{y} \\
      L_{y'}(x,y(x),y'(x)) &= \od{f(x)}{y'} \\
    \end{align*}
  \item Use an Euler equation and solve for $y(x)$. Invoke boundary conditions
    to resolve the constants. This is the extremal.
  \end{enumerate}
\item[1st Order Fixed-Free] Given
  $$F(y) = \int_a^b f(x)\dd{x}, \quad y(a)=x_0, \quad y(b)\;\text{free}$$
  \begin{enumerate}
  \item Let
    \begin{align*}
      L(x,y(x),y'(x)) &= f(x) \\
      L_{y}(x,y(x),y'(x)) &= \od{f(x)}{y} \\
      L_{y'}(x,y(x),y'(x)) &= \od{f(x)}{y'} \\
    \end{align*}
  \item Use an Euler equation and solve for $y(x)$. Invoke the boundary
    condition to resolve a constant. You will have one unresolved constant after
    this.
  \item Invoke the Natural Boundary Condition $L_{y'}=0$ about the free
    endpoint. This will give a second boundary.
  \item Use this second boundary to resolve the remaining constant.
  \end{enumerate}

{\tiny
Find the extremal for
\begin{equation}
  \label{eq:5a-problem}
  \begin{aligned}
    J(y) = \int_0^3 e^{2x}(y'^2-y^2)\dd{x}, \\
    y(0)=1, y(3)=\text{free}. \\
  \end{aligned}
\end{equation}
\begin{align*}
  L(x,y,y') &= e^{2 x} \left({\left(y'\right)}^2-y^2\right) \\
  L_y(x,y,y') &= -2 e^{2 x} y \\
  L_{y'}(x,y,y') &= 2 e^{2 x} y' \\
\end{align*}
\begin{equation}
\label{eq:5a-ode}
\begin{aligned}
  0 &= -2 e^{2 x} y - \od{}{x}2 e^{2 x} y' \\
    &= -2 e^{2 x} y - 4e^{2x}y' - 2e^{2x}y'' \\
  0 &= y + 2y' + y''
\end{aligned}
\end{equation}
Based on the non-free condition, this ODE may be solved by standard methods
\begin{equation*}
  \implies y(x) = e^{-x}(1+c_2x).
\end{equation*}
The remaining constant $c_2$ may be determined by natural boundary condition
\begin{align*}
  L_{y'}(b,y(b),y'(b)) &= 2 e^{6} y'(3) \\
  0 &= 2e^{6}y'(3) \\
  \implies y'(3) &= 0.
\end{align*}
We may re-solve by standard methods with this additional
condition:
\begin{equation*}
  \boxed{y(x) = -\frac{1}{2} e^{-x} (x-2).}
\end{equation*}
}

\item[2nd Order Fixed-Fixed] Use the same method for 1st order fixed-fixed, just
  involving another term in the Lagrangian.

{\tiny
Find the extremals for the functional:
\begin{equation}
  \label{eq:4.4.1d-problem}
  \begin{aligned}
    J(y) &= \int_0^1 (yy'+{(y'')}^2)\dd{t} \\
    y(0) &= 0, \quad y(1) = 2, \quad y'(0) = 1,\quad y'(1) = 4 \\
  \end{aligned}
\end{equation}
\begin{align*}
  L &= yy'+{(y'')}^2, \\
  L_y &= y', \\
  L_{y'} &= y, \\
  L_{y''} &= 2y''. \\
\end{align*}
\begin{align*}
  0 &= L_y - \od{}{x}L_{y'} + \od[2]{}{x}L_{y''} \\
    &= y' - \od{}{x}y + \od[2]{}{x}2y'' \\
    &= y'''' \\
\end{align*}
\begin{equation*}
  \boxed{y(x) = x + x^3}
\end{equation*}
}

\item[Multivariable Fixed-Fixed] Same method as fixed-fixed, just have two
  equations. Solve for each $y_i$ independently.
